No Evidence of Bias in Arrests Made by the Toronto Police Service

### 2024-01-23 outputs/paper/paper.qmd
Filter Updates
- Updated Cleaned Data
- Added Scatter Plot for Results

---
title: "No Evidence of Bias in Arrests Made by the Toronto Police Service"
author: "Aviral Bhardwaj"
abstract: The ethnic identity of citizens arrested by the police has consistently been a topic of discussion, often highlighting the pronounced skewness of certain ethnic identities. This paper seeks to analyze data from the Toronto Open Data Portal to investigate whether there is any bias in arrests made by the Toronto Police Service between 2020 and 2021, while accounting for factors such as age, booked after arrest, and search on arrest. While the population distribution of various ethnic identities varies significantly, it was found that ..... Further analysis can provide valuable insights for policymakers, law enforcement agencies, and community advocates in fostering a more equitable society while emphasizing the importance of promoting transparency and accountability in policing.
date: today
date-format: long
number-sections: true
format: pdf
bibliography: references.bib
---

@@

# Mutating the date for age groups involving Aged 17 years and younger and Aged 17 years and under essentially they mean the same thing
cleaned_data <- cleaned_data %>%
  mutate(age_group = case_when(
    age_group == "Aged 17 years and younger" ~ "Under 18 years",
    age_group == "Aged 17 years and under" ~ "Under 18 years",
    TRUE ~ age_group
  ))

@@

table2_long <-
  cleaned_data |>
  select(
    arrest_year,
    perceived_race,
    booked
  ) |>
  group_by(arrest_year, perceived_race, booked) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(perceived_race)) %>%
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

@@

### Number of Arrests by Age Group for the year 2020-2021
As seen in Figure 3, the data indicates a non-uniform distribution of arrests across age groups, with the age group of 25 to 34 years and age group 35 to 44 exhibiting the highest number of arrests in both 2020 and 2021. Furthermore, the data indicates a decrease in the number of arrests for the top two age groups from 2020 to 2021. However, a contrasting trend emerges for other age groups, where the number of arrests has increased over the same time period.

@@

table3_age_24_long <- pivot_longer(table3_small, cols = c("2020", "2021"), names_to = "year", values_to = "value")

age_35_44 <- cleaned_data %>%
  filter(age_group == "Aged 35 to 44 years") |>
  group_by(arrest_year, perceived_race, age_group) %>%
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0) |>
  pivot_longer(cols = c("2020", "2021"), names_to = "year", values_to = "value")

# Combining the data
table3_age_24_long <- rbind(table3_age_24_long, age_35_44)

ggplot(table3_age_24_long, aes(x = perceived_race, y = value, fill = factor(year))) +
  geom_bar(stat = "identity", position = 'dodge') +
  facet_wrap(~ age_group, scales = "free_y") +
  labs(title = "Bar Chart for the number of arrests by Age-Group for the year 2020, and 2021",
       x = "Perceived Race",
       y = "Value",
       fill = "Age Group") +
  theme_minimal() +
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## Discussion
## References

>>>>

---
title: "No Evidence of Bias in Arrests Made by the Toronto Police Service"
author: "Aviral Bhardwaj"
abstract: The ethnic identity of citizens arrested by the police has consistently been a topic of discussion, often highlighting the pronounced skewness of certain ethnic identities. This paper seeks to analyze data from the Toronto Open Data Portal to investigate whether there is any bias in arrests made by the Toronto Police Service between 2020 and 2021, while accounting for factors such as age, booked after arrest, and search on arrest. While the population distribution of various ethnic identities varies significantly, it was found that ..... Further analysis can provide valuable insights for policymakers, law enforcement agencies, and community advocates in fostering a more equitable society while emphasizing the importance of promoting transparency and accountability in policing.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
bibliography: references.bib
---

@@

# Mutating the date for age groups involving Aged 17 years and younger and Aged 17 years and under essentially they mean the same thing
cleaned_data <- cleaned_data %>%
  mutate(age_group = case_when(
    age_group == "Aged 17 years and younger" ~ "Aged 17 years and under",
    age_group == "Aged 17 years and under" ~ "Aged 17 years and under",
    TRUE ~ age_group
  ))

@@

table2_long <-
  cleaned_data |>
  select(
    arrest_year,
    perceived_race,
    booked
  ) |>
  group_by(arrest_year, perceived_race, booked) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(perceived_race)) %>%
  filter(!(perceived_race == "None")) %>%
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

@@

### Number of Arrests by Age Group for the year 2020-2021
As depicted in Figure 3, the data reveals a non-uniform distribution of arrests across age groups, with the age groups of 25 to 34 years and 35 to 44 years exhibiting the highest number of arrests in both 2020 and 2021. Furthermore, there is a notable decrease in the number of arrests for these top two age groups from 2020 to 2021. In contrast, a different trend emerges for other age groups, where the number of arrests has increased over the same time period.

Examining Figure 4, which illustrates the dot plot for the number of arrests by age group in the year 2020 and 2021 for the top two ethnic groups, "Black" and "White," it becomes evident that the age group 25 to 34 years represents the highest number of arrests for individuals of Black ethnicity. The trend for Black ethnicity indicates a decreasing curve from this age group onward. In contrast, for individuals of White ethnicity, the peak in arrests is observed in the age group 35 to 44 years. It is noteworthy that from age 45 to 65, the difference in the number of arrests between both ethnicities starts to converge.

@@

table3_ii <- table3 %>%
  filter(perceived_race %in% c("White", "Black")) %>%
  filter(!(age_group == "None")) %>%
  pivot_longer(cols = c("2020", "2021"), names_to = "year", values_to = "value")

# Create a line Graph
ggplot(table3_ii, aes(x = age_group, y = value, color = perceived_race)) +
  geom_point() + 
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ year, scales = "free_y") +
  labs(title = "Scatter Plot for the number of arrests by Age-Group for the year 2021",
       x = "Age Group",
       y = "Value",
       color = "Perceived Race") +  # Change fill to color
  theme_minimal() +
  theme(legend.position = "right") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
## Discussion

## References
